---
title: 'Simple chi calculations'
date: "`r Sys.Date()`"
author: 'Vincent Godard'
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: yes
    df_print: paged
    code_download: true
    code_folding: show
---

# Objectives

This tutorial demonstrates simple cases of $\chi$ calculation with the `gtbox` package.

- computing a $\chi$  map


The first thing we have to do is to load the `gtbox` package (once it has been installed).
It is required to have GRASS GIS (version 7) installed, as some of the functions used here will rely heavily on GRASS raster processing modules (through the `rgrass7` package).
It will also be necessary to install the following Addons (through the command `g.extension` in a GRASS session) :

- `r.stream.distance`
- `r.stream.basins`
- `r.stream.order`

Note that we will use the `terra` package for most geospatial operations in R. We will use the standard formats from this package : `SpatRaster` for raster layers (eventually with multiple layers) and `SpatVector` for vector layers.

More information on the terra package can be found here :

- [https://rspatial.org/terra/index.html](https://rspatial.org/terra/index.html)

```{r message=FALSE, warning=FALSE}
library("gtbox")
library("terra")
```



# Preliminary steps

We first load a DEM.

```{r}
data("dem_cevennes",package = "gtbox")
dem = terra::rast(dem_cevennes)
rm(dem_cevennes)
```

We can compute a hillshade surface and use some transparency on the DEM for a nicer display.

```{r}
slope <- terra::terrain(dem, "slope", unit="radians")
aspect <- terra::terrain(dem, "aspect", unit="radians")
hill <- terra::shade(slope, aspect, 40, 270)
plot(hill, col=grey(0:100/100), legend=FALSE)
plot(dem, col=rainbow(25, alpha=0.35), add=TRUE)
```

We then process the DEM, as illustrated in the Simple Digital Elevation Model operations vignette.

```{r warning=FALSE}
gisbase = "/usr/lib/grass78/"
th_px = 2000 # stream initiation threshold in pixels
st = process_dem(dem,th_px,gisbase=gisbase)
```

# $\chi$ map computation

We are going to compute a chi map over the DEM, starting the integration at a specified elevation.
We first get the corresponding outlets and basins.

```{r}
outlets = get_outlets(st,elevation=250,gisbase=gisbase)
basins = get_basins(st,outlets,gisbase=gisbase)
```

We then compute the $\chi$ values along the network.

```{r}
chi = compute_chi(st,outlets,a0=1000,mn=0.5,gisbase)
```

We transfer the $\chi$ values defined along streams to adjacent areas to obtain a continuous map, easier to visualize.

```{r}
chi_map = streams2basins(st,chi,FUN="mean")
chi_map = mask(chi_map,basins)
```

We plot everything

```{r}
plot(hill, col=grey(0:100/100), legend=FALSE)
plot(chi_map, col=heat.colors(25, alpha=0.35), add=TRUE)
lines(basins)
points(outlets)
```









































