---
title: 'Swath profile calculation '
date: "`r Sys.Date()`"
author: 'Vincent Godard'
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: yes
    df_print: paged
    code_folding: show
    code_download: true
---

```{r include=FALSE}
library(DT)
```


We load a number of packages that will be used 


```{r message=FALSE, warning=FALSE}
library("geomorphology")
library("raster")
library("rgdal")
library("sp")
library("rgeos")

```

We load and reproject a Digital Elevation Model over the Cévennes area in SE France.
```{r}
data("dem_cevennes",package = "geomorphology")
dem = dem_cevennes
rm(dem_cevennes)
#dem0 = raster("srtm_cevennes.tif") # we could import our own DEM this way
dem <- projectRaster(dem, crs="EPSG:32631",res=30,method="bilinear") # project to UTM 31N
dem = trim_na(dem) # we trim to get a nice rectangular DEM with no NA on the sides
```


First we can read a vector file containing the trace of the profile, define a width `W` and plot the corresponding buffer.

```{r }
data("profile_cevennes") # importing the profile
profile = profile_cevennes
rm(profile_cevennes)
#profile = readOGR("../data/profile_cevennes.gpkg",verbose=FALSE) # we could import our own data this way
W = 5e3 # define the one sided width of the swath in m
plot(dem,col=terrain.colors(100,rev = FALSE))
lines(profile,lwd=2)
lines(gBuffer(profile,width=W,capStyle = "FLAT")) # plot a buffer on the map
```

Then we extract the start and end points of the profile from the line (we could also define them manually).
We will use `inc` bins along the profile.

```{r }
x1 =  profile@lines[[1]]@Lines[[1]]@coords[1,1]
y1 =  profile@lines[[1]]@Lines[[1]]@coords[1,2]
x2 =  profile@lines[[1]]@Lines[[1]]@coords[2,1]
y2 =  profile@lines[[1]]@Lines[[1]]@coords[2,2]
inc = 100 # number of bins along the profile
data = swath_profile(dem,x1,y1,x2,y2,W,100) # computation of the swath profile
```

The output is the following table with `r inc` rows, containing the basic statistics for each bin.

```{r}
datatable(data)
```


We can then plot the profile.

```{r}
plot(data$distance/1000,data$mean,type="l",lwd=3,ylim=range(data$min,data$max),xlab="Distance along profile (km)",ylab="Elevation (m)")
lines(data$distance/1000,data$mean+data$sd)
lines(data$distance/1000,data$mean-data$sd)
lines(data$distance/1000,data$min,lty=2)
lines(data$distance/1000,data$max,lty=2)
```



