---
title: 'Swath profile calculation '
date: "`r Sys.Date()`"
author: 'Vincent Godard'
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: yes
    df_print: paged
    code_folding: show
    code_download: true
---

```{r include=FALSE}
library(DT)
```


We load a number of packages that will be used 


```{r message=FALSE, warning=FALSE}
library("gtbox")
library("terra")
library("sf")
```

We load and reproject a Digital Elevation Model over the Cévennes area in SE France.
```{r}
data("dem_cevennes",package = "gtbox")
dem = terra::rast(dem_cevennes)
rm(dem_cevennes)
plot(dem)
```


First we can define the trace of the profile (using the coordinates of the start and end points), the width `w_buf` and plot the corresponding buffer.

```{r }
# start (1) and end (2) points of the projection line
x1 = 583695
y1 = 4875921
x2 = 550939
y2 = 4907255

w_buf = 5e3 # width 
# projection line
l1 = terra::vect(matrix(c(x1,y1,x2,y2),ncol =2,byrow=T),
                 type="lines",crs=crs(dem))
# swath 
bf = terra::vect(st_buffer(st_as_sf(l1),dist=w_buf,
                    endCapStyle = "FLAT",joinStyle = "ROUND"))
plot(dem)
lines(l1)
lines(bf,lty=2)
```

Then we project the DEM data along the profile, using `swath_profile` function.
We will use `inc` bins along the profile.

```{r }
inc = 100 # number of bins along the profile
data = swath_profile(dem,x1,y1,x2,y2,w_buf,100) # computation of the swath profile
```

The output is the following table with `r inc` rows, containing the basic statistics for each bin.

```{r}
datatable(data)
```


We can then plot the profile.

```{r}
plot(data$distance/1000,data$mean,type="l",lwd=3,ylim=range(data$min,data$max),xlab="Distance along profile (km)",ylab="Elevation (m)")
lines(data$distance/1000,data$mean+data$sd)
lines(data$distance/1000,data$mean-data$sd)
lines(data$distance/1000,data$min,lty=2)
lines(data$distance/1000,data$max,lty=2)
```



