---
title: 'Stochastic incision model '
date: "`r Sys.Date()`"
author: 'Vincent Godard'
bibliography: '`r system.file("REFERENCES.bib",package = "geomorphology") `'
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: yes
    df_print: paged
    code_folding: show
    code_download: true
---

We load the `geomorphology` package. 

```{r message=FALSE, warning=FALSE}
library("geomorphology")
```

The theoretical background for the incorporation of discharge probability distributions into rivr incision models have been extensively described in previous studies (@tucker2000stochastic, @lague2005discharge, @dibiase2011influence, @scherler2017testing, @campforts2020parameterization), and its main features will be presented briefly below.

# Discharge distribution

In incision models taking into account discharge variability, the probability distribution of discharges is often modelled using an inverse gamma law,
\begin{equation}\label{eq:inv_gamma}
pdf(Q^*) = \frac{k^{k+1}}{\Gamma(k+1)} \exp(-\frac{k}{Q^*})  Q^{*-(k+2)},
\end{equation}
where $Q^* = Q/\bar{Q}$ is normalized discharge, $k$ is the variability parameter, and $\Gamma$ is the Gamma function.
Low values of $k$ correspond to situation where the discharge variability is high.
The inverse gamma distribution combines exponential and power law parts describing low and high discharge regimes, respectively, and accounting for the low probabilities of events at both ends of the spectrum. 

We plot below a few examples using the `Inv_Gamma_dist` function. 
In this case we considered normalized discharge, so $\bar{Q} = 1$.

```{r}
par(mar = c(3,3,1,1)) # margins size
Qs=10^seq(-1.3,3,length.out = 100) # vector of normalized discharge
plot(NA,log="xy",xlim=range(Qs),ylim=c(1e-8,5),yaxs="i",axes=F, xlab=NA, ylab=NA)
axis(side = 2,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8,at=10^c(-8,-6,-4,-2,0),labels=expression(10^-8,10^-6,10^-4,10^-2,10^0))
mtext(side = 2, line = 1.5, "Probability",cex=1)
axis(side = 1,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8,at=10^c(-1,0,1,2,3),labels=expression(10^-1,10^0,10^1,10^2,10^3))
mtext(side = 1, line = 1.25, "Normalized Discharge  (-)",cex=1)
abline(v=1)
Ig = Inv_Gamma_dist(0.5,Qs,1)
lines(Qs,Ig,lwd=3)
Ig = Inv_Gamma_dist(1,Qs,1)
lines(Qs,Ig)
Ig = Inv_Gamma_dist(0.25,Qs,1)
lines(Qs,Ig,lty=3)
Ig = Inv_Gamma_dist(5,Qs,1)
lines(Qs,Ig,lty=2)
Ig = Inv_Gamma_dist(20,Qs,1)
lines(Qs,Ig,lty=2,col="grey")
legend("topright",c("k=0.25","k=0.5","k=1","k=5","k=20"),lty=c(3,1,1,2,2),lwd=c(1,3,1,1,1),col=c(rep("black",4),"grey"))
box()
```

The previous figure show the evolution of the shape of the inverse gamma distribution with the variability parameter `k`.

# Incision model

## Instantaneous incision

Instantaneous detachment-limited fluvial incision $I$ can be expressed as a function of shear stress acting on the river bed $\tau$,
\begin{equation}
\label{eq:inst_inc_1}
I = k_e \left( \tau^a  - \tau_c^a \right),
\end{equation}
where $\tau_c$ is a threshold shear stress beyond which incision occurs, $k_e$ is an erodibility coefficient depending on the properties of the bedrock, and $a$ an exponent dependent on the nature of incision processes.

Shear stress $\tau$ can be in turn expressed as a function of discharge $Q$, slope $S$ and channel width $W$ as, 
\begin{equation}\label{eq:tau}
\tau = k_t \left( \frac{Q}{W} \right)^\alpha S^\beta,
\end{equation}
with $k_t$, $\alpha$ and $\beta$ parameters depending on the formulation used to express flow resistance.
For example, $\alpha=3/5$ and $\beta=7/10$ when using Manning's frictional relationship (@howard1994detachmentlimited).

The variation of channel width $W$ downstream and with instantaneous discharge $Q$ can be computed according to,
\begin{equation}
\frac{W}{W_b} = \left( \frac{Q}{\bar{Q}}  \right)^{\omega_s},
\end{equation}
with $\omega_s$ an empirical parameter and $W_b$ the bankfull discharge, scaling with the mean annual discharge $\bar{Q}$ as,
\begin{equation}\label{eq:wb}
W_b = k_w \bar{Q}^{\omega_b},
\end{equation}
where $k_w$ and $\omega_b$ are another couple of empirical parameters, which can be determined from field measurements or remote sensing imagery @kirby2011tectonic @dibiase2011influence @fisher2012channel.

Mean annual discharge $\bar{Q}$ can be calculated from mean runoff $\bar{R}$ and drainage area $A$ as $\bar{Q} = \bar{R}A^c$, with $c$ an exponent that we will consider equal to 1 in the following.

Equations above can then be combined to yield an expression for instantaneous incision $I$.
\begin{equation}\label{eq:inst_inc_2}
I = k_e k_t^a k_w^{-a\alpha}\bar{R}^{a\alpha(1 - \omega_b)}  \left(\frac{Q}{\bar{Q}}\right)^{a\alpha(1-\omega_s)} A^{a\alpha(1-\omega_b)}S^{a\beta} - k_e\tau^a
\end{equation}

This equation can be simplified by defining an area exponent $m=a\alpha(1-\omega_b)$, a slope exponent $n=a\beta$, an erodibility term $K=k_e k_t k_w^{-a\alpha}\bar{R}^m$, $\gamma = a\alpha(1-\omega_s)$ and a threshold term $\Psi=k_e\tau^a$, which yields,
\begin{equation}\label{eq:inst_inc_3}
I= K \left(\frac{Q}{\bar{Q}}\right)^\gamma A^m S^n - \Psi
\end{equation}
This relationship highlight the importance of the parameter $\gamma$ in controlling the sensitivity of incision to discharge and its variability (@lague2005discharge).

Introducing the steepness index $k_s  = A S^{m/n}$ (@kirby2012expression), the previous equation  can also be written as,
\begin{equation}\label{eq:inst_inc_3_ksn}
I= K \left(\frac{Q}{\bar{Q}}\right)^\gamma k_s^n - \Psi
\end{equation}
which can be computed with function `I_instantaneous`.

If we assume that the threshold term $\Psi$ is negligible and that discharge $Q$ is constant and equal to mean discharge $\bar{Q}$, we can simplify to the classical stream power form,
\begin{equation}\label{eq:inst_inc_cst_q}
I= K  A^m  S^n
\end{equation}

## Long-term incision

Long term incision $E$ can be calculated by weighting instantaneous incision  by discharge probability $pdf(Q)$ and integrating over $Q$ .
\begin{equation}\label{eq:long_inc}
E = \int_{Q_c}^{Q_m} I(Q,k_s) pdf(Q) dQ,
\end{equation}
where $Q_c$ is the critical discharge needed to overcome the incision threshold and $Q_m$ is the maximum discharge.
$Q_c$ is obtained by setting $I=0$ in the instantaneous incision equation,
\begin{equation}\label{eq:qc}
Q_c = \bar{Q} \left( \frac{\Psi}{K} \right)^{\frac{1}{\gamma}} k_s^{-\frac{n}{\gamma}},
\end{equation}	
and is calculated by function `Qcritical`.


# Relationship between channel steepness and long-term incision

We can now compute the predicted long-term incision rate $E$ for a given channel steepness and a value o the variability parameters $k$.

We use the same parameters as @dibiase2011influence, which are defined below.

```{r}
# Table 2 from DiBiase and Whipple (2011)
ke = 4.3e-12 # erodibility parameter  
tauc = 45 # critical shear stress (Pa)
kq = 9e-9 # !!! not used?
kw = 15 # empirical constant for downstream variation in channel width (equation 4a)
kt = 1000 # equation 3
omegas = 0.25 # empirical exponent for at-a-station variation in channel width (equation 4b)
omegab = 0.55 # empirical exponent for downstream variation in channel width (equation 4a)
a = 3/2   # equation 2 shear stress exponent
alpha = 2/3 # equation 3
beta = 2/3 # equation 3
c = 1 # equation 6
Rb = 0.28/(365*24*3600)  # mean daily run off (m/s) (equation 6)

# derived parameters for equation 7
m = alpha*a*(1-omegab)
n = beta*a

# derived parameters for equation 8
K = ke*kt^a*kw^(-1*alpha*a)*Rb^m
gamma = alpha*a*(1-omegas)
Psi = ke*tauc^a
```

The calculation is done by the `I_long_term` function.


```{r}
# plot definition
par(mar = c(3,3,1,1)) # margins size
plot(NA,xlim=c(0.1,1200),ylim=c(0,300),yaxs="i",xaxs="i",axes=F, xlab=NA, ylab=NA)
axis(side = 1,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8)
mtext(side = 1, line = 1.5, "Erosion rate (m/Ma)",cex=1)
axis(side = 2,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8)
mtext(side = 2, line = 1.25, "Steepness index",cex=1)
grid()
#
ks = seq(1,300,length.out = 20) # define the values of ks for which we wnat to compute.
df = data.frame(k=c(2,1,0.7,0.5,0.35,0.2),lty=c(2,2,2,1,3,3),col=c("grey85","grey50","black","black","grey50","grey85")) 
df$leg = paste("k =",df$k)
#
for (i in 1:nrow(df)){
  E = I_long_term(df$k[i],ks,Psi,K,n,gamma)*(365*24*3600)*1e6
lines(E,ks,lty=df$lty[i],lwd=3,col=as.character(df$col[i]))
}
#
box()
legend("bottomright",df$leg,lty=df$lty,lwd=3,col=as.character(df$col),ncol=2)

```

This is basically figure 7b from @dibiase2011influence.
We can also try to change run off $R_b$.

```{r}
# plot definition
par(mar = c(3,3,1,1)) # margins size
plot(NA,xlim=c(0.1,1200),ylim=c(0,300),yaxs="i",xaxs="i",axes=F, xlab=NA, ylab=NA)
axis(side = 1,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8)
mtext(side = 1, line = 1.5, "Erosion rate (m/Ma)",cex=1)
axis(side = 2,tcl=-0.25,mgp=c(1,0.25,0),cex.axis=0.8)
mtext(side = 2, line = 1.25, "Steepness index",cex=1)
grid()
#
k = 0.5
ks = seq(1,300,length.out = 20) # define the values of ks for which we want to compute.
df = data.frame(Rb=c(10,5,2,1,0.5,0.2,0.1),lty=c(2,2,2,1,3,3,3),col=c("grey85","grey50","grey25","black","grey25","grey50","grey85"))
df$leg = paste("Rb x",df$Rb)
#
for (i in 1:nrow(df)){
  K = ke*kt^a*kw^(-1*alpha*a)*(df$Rb[i]*Rb)^m
  E = I_long_term(k,ks,Psi,K,n,gamma)*(365*24*3600)*1e6
lines(E,ks,lty=df$lty[i],lwd=3,col=as.character(df$col[i]))
}
#
box()
legend("bottomright",df$leg,lty=df$lty,lwd=3,col=as.character(df$col),ncol=2)

```




# References




